---
title: "campR -- ggplot2"
author: "[Marcus Vollmer](https://github.com/MarcusVollmer/campR)"
date: "2019-01-17"
output:
  html_document:
    toc: true
    toc_depth: 3
---

This is a [campR](https://github.com/MarcusVollmer/campR) notebook introducing ggplot2.

## Libraries and data

The following commands imports the libraries of **ggplot2** and **ggplot2movies**.
The latter package makes the data frame *movies* available a collection of motion pictures with user ratings frome the internet movie database (IMDB). We take a short look on the variables available.
```{r}
library(ggplot2)
library(ggplot2movies)
head(movies)

```

## Basic graphics building
The [ggplot2 cheat sheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf) summarizes the programming syntax.

> ggplot(data = \<DATA\>) +  
> \<GEOM_FUNCTION\>(mapping=aes(\<MAPPINGS\>), stat=\<STAT\> , position=\<POSITION\>) +  
> \<COORDINATE_FUNCTION\> +  
> \<FACET_FUNCTION\> +  
> \<SCALE_FUNCTION\> +  
> \<THEME_FUNCTION\>


```{r echo=TRUE}
# Lets define the data basis
  (p0 = ggplot(data=movies, aes(year)))

# and add a geom layer to draw a histogram
  (p1 = p0 + geom_histogram())

# We can change the breaks of the histogram in this way
  (p1 = p0 + geom_histogram(breaks=seq(1900,2010,by=5)))

# We add a title and rename the x-axis
  (p1 = p1 + labs(title='Movie distribution', subtitle='Imdb listed movies', x='Release date'))
```


### Geoms
Geoms describes the type of geometric objects that represent data.
```{r echo=TRUE}
# We add another geometrical layer to the graph which smooths the discrete density
  p0 + geom_histogram(aes(y=..density..)) + geom_density(alpha=.2, fill="#FF6666")

# The histogram can be separated according to a factor variable
  ggplot(data=movies, aes(year, color=factor(Animation))) +
    geom_histogram()
``` 


### Scales
Scales describes how visual characteristics are converted to displayed values.
```{r echo=TRUE}
# Lets show the counts on log scale
  p1 + scale_y_log10()
```
  
### Facets
Facets describes how data is split into subsets and displayed as multiple graphs.
```{r echo=TRUE} 
# To show the data in panels use facets
  (p2 = ggplot(data=movies, aes(year, color=factor(Animation))) +
    geom_histogram(aes(y=..density..)) + 
    facet_grid(factor(Animation) ~ .) + 
    geom_density(alpha=0.6))
  
# You can also define a grid of facets
  p2 + facet_grid(factor(Animation) ~ factor(Action))
  
# Append the facet grid to use three variables
  p2 + facet_grid(factor(Animation) ~ factor(Action) + factor(Comedy))

# Confused about the binary labelling? So change the facet labels from {0,1} to self-descriptive labels.
# Labeling the factor level works fine:
  movies$Action = factor(movies$Action, c(0,1), c('No action', 'Action'))
  ggplot(data=movies, aes(year, color=factor(Animation))) +
    geom_histogram(aes(y=..density..)) +
    geom_density(alpha=0.6) +
    facet_grid(factor(Animation) ~ Action) 
```

### Themes
Themes describes the look of non-data components of the plot.
Get informed on named colors in R on http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
```{r echo=TRUE} 
# Add a theme layer to graph
  p2 + theme_classic()
  p2 + theme_bw()
  p2 + theme_dark()
  p2 + theme_void()

# or changing various properties of the theme
  p2 + theme(axis.ticks = element_line(size = 2),
             panel.background = element_rect(fill='cornsilk2'))
  
# you can also use the corresponding hex color or RGB color codes:
  p2 + theme(axis.ticks = element_line(size = 2),
             panel.background = element_rect(fill='#eee8cd')) 
  p2 + theme(axis.ticks = element_line(size = 2),
             panel.background = element_rect(fill=rgb(238/255,232/255,205/255))) 
  
  
# Themes can also handle legends
  p2 + theme(legend.justification = "top")
  p2 + theme(legend.position = c(.25, .95),
    legend.justification = c("right", "top"),
    legend.margin = margin(6, 6, 6, 6))
```


## Scatterplot
```{r}
# Lets plot the movie budget together with the average IMDB rating
  (p3 = ggplot(movies, aes(budget,rating)) +
    geom_point())

# Mark the most prominents movies
  library(ggrepel)
  p3 + geom_label_repel(data=subset(movies, budget>1.5*10^8), aes(label=title, fill=factor(length<120)), size=3.5, box.padding=unit(0.35, "lines"), point.padding=unit(0.3, "lines")) +
    theme(legend.position="bottom")

# Lets rely on minimum 100 votes and use a log scale
  (p4 = ggplot(subset(movies, votes>100), aes(budget,rating)) +
    geom_point() +
    scale_x_continuous(trans='log10'))
  
# Choose hexagonal geometry to summarize/cluster multiple data points
  library(hexbin)
  p4 + geom_hex()
```

... to be continued ...