---
title: "campR -- summary tables"
author: '[Marcus Vollmer](https://github.com/MarcusVollmer/campR)'
institute: "Institut of Bioinformatics, University Medicine Greifswald"
date: '2019-06-06'
header-includes: 
   - \usepackage{rotating}
   - \usepackage{longtable}
output:
  html_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: '3'
    latex_engine: xelatex
---

```{r setup, include=FALSE}
is_html = function() {knitr::opts_knit$get("rmarkdown.pandoc.to")=="html"}
is_latex = function() {knitr::opts_knit$get("rmarkdown.pandoc.to")=="latex"}

library(knitr)
opts_chunk$set(tidy=TRUE)
```


This is a [campR](https://github.com/MarcusVollmer/campR) notebook introduces summary tables with 'xtable', 'stargazer', 'rapport' and show how to use own helper functions for group comparisons.

Rapport needs 'Pandoc' which is not available in all R versions. After that run: 'install_github(\'rapport\', \'rapporter\')'.

# Summary tables {.tabset .tabset-fade .tabset-pills}
## Libraries and data

The following commands imports the libraries of **xtable**, **stargazer** and **rapport**. Further, we import some user defined function from GitHub.
The latter package makes the data frame *movies* available a collection of motion pictures with user ratings frome the internet movie database (IMDB).
```{r}
library(xtable)
library(stargazer)
library(rapport)

# Load user created functions
  library(devtools)
  repo = "https://raw.githubusercontent.com/MarcusVollmer/R-Helper-Functions/master/"
  source_url(paste0(repo,"strtable.R"))
  source_url(paste0(repo,"stargazer_long.R"))
  source_url(paste0(repo,"characteristics_table.R"))
  source_url(paste0(repo,"mytable.R"))

library(ggplot2movies)
```

\newpage
## xtable

Lets have short look what kind of data xtable can process:
```{r}
methods(xtable)
```

The [xtable documentation](https://cran.r-project.org/web/packages/xtable/xtable.pdf) summarizes the programming syntax and lists available functions.

We will plot the head of the data and will display the summary of a linear model in the default R:markdown fashion.
```{r echo=TRUE}
# Lets quickly check the movie data base
  (h = head(movies))
  movies$genre = factor(with(movies, paste(Action, Comedy, Drama, Documentary, Romance, Short, sep="")), c("100000","010000","001000","000100","000010","000001"), c("Action", "Comedy", "Drama", "Documentary", "Romance", "Short"))

# Build a generalized linear model to explain the budget
  mymodel = lm(budget ~ length + rating + year + genre, movies)
  (s = summary(mymodel))
```
 
Next, we will plot the results with **xtable** as \LaTeX and HTML output.
```{r echo=TRUE, results='asis', include=is_latex()}
# The basic LaTeX output
  xtable(h)
  xtable(mymodel)
```

```{r echo=TRUE, results='asis', include=is_html()}
# The basic HTML output
  print(xtable(h), type="html")
  print(xtable(mymodel), type="html")
        
# Alternatively use kable for html tables
  knitr::kable(h, caption="Table with kable")
  
# This results in an error
  #knitr::kable(mymodel, caption="Table with kable")
  # Error in as.data.frame.default(x) : cannot coerce class ""lm"" to a data.frame
```


If you need to change some properties use **options**:
```{r echo=TRUE, results='asis', include=is_html()}
# Changing default options for the entire markdown file are quite easy
  options(xtable.caption="Nice table")
  options(xtable.timestamp="")
  
# To see the options to set **print.xtable**:
  options(xtable.rotate.colnames=TRUE)
  options(xtable.include.rownames=FALSE)
  print(xtable(h), type="html")
```

```{r echo=TRUE, results='asis', include=is_latex()}
# The basic LaTeX output
  xtable(h)
  xtable(mymodel)
```

\newpage
## stargazer

Stargazer produces LaTeX code, HTML/CSS code and ASCII text for well-formatted tables.
```{r echo=TRUE, results='asis', include=is_latex()}
# The basic LaTeX output
  stargazer(h)
  stargazer(mymodel)
  
#
  stargazer(h, summary=FALSE)
```

Lets have short look at the supported objects:
```{r echo=TRUE}
  ?`stargazer models`
```

Stargazer in action to display results of linear models:
```{r echo=TRUE, results='asis', include=is_latex()}
# Lets model the budget of movies with and without the interaction of movie length and genre
  m1 = lm(budget ~ length + genre + rating + year, movies)
  m2 = lm(budget ~ length*genre + rating + year, movies)
  
# Stargazer has a nice combined representation of multiple models
  stargazer(m1, m2, title="Regression results")
  
# You have various properties you can change according to your need
  stargazer(m1, m2, title="Regression results",
            single.row=TRUE, ci=TRUE, ci.level=0.95,
            decimal.mark='.', digits=0)
```

Or you can use the html output:
```{r echo=TRUE, results='asis', include=is_html()}
# Stargazer has a nice combined representation of multiple models
  stargazer(m1, m2, type="html", title="Regression results")
  
# You have various properties you can change according to your need
  stargazer(m1, m2, type="html", title="Regression results",
            single.row=TRUE, ci=TRUE, ci.level=0.95,
            decimal.mark='.', digits=0)
```


\newpage
## Own summary functions

<R-Helper-Functions>[https://github.com/MarcusVollmer/R-Helper-Functions](R-Helper-Functions) can be sourced directly from GitHub to fasten and to standardize the analysis and to allow reproducability.

Currently there are four useful functions provided:

1. strtable.R -- a modified str implementation for data frames (originally published by Jason Bryer)
2. stargazer_long.R -- a modification of stargazer to display long tables, to allow rotations and LaTeX column specifications
3. mytable.R -- a cross-table for factor variables to display counts or summary statistics of a continuous variable with univariable statistical testing
4. characteristics_table.R -- creates stratified tables of data along with univariable statistical testing


### strtable and stargazer_long

```{r, results='asis', include=is_latex()}
# Some genre columns should be set to logicals
  booleanvars = colnames(movies[,18:24])
  movies[,booleanvars] = lapply(booleanvars, function(x) as.logical(movies[,x]))

# Generate a summary table using strtable
  s = strtable(movies, n=10, width=300)

# For LaTeX output as a longtable use the follwing command.
# The compiling of the LaTeX script to PDF need the inclusion of the **longtable** in the header.
  stargazer_long(s[, 1:4], summary=FALSE, rownames=FALSE, output="cccp{9cm}", rotate=60)
```

### mytable
#### Cross table for count data with confidence intervals

Printing a cross table with p value and percentages (rows sums up to 100%).
```{r, results='asis', include=is_latex()}
  library(PropCIs)
  mytable(movies$Animation, movies$Action, ci=FALSE, prec="%.2f", latex=TRUE)
```

Printing the same table with 95% exact Clopper-Pearson confidence intervals (PropCIs package required) and less digits with % sign, a fixed column width and floating point p-value style:
```{r, results='asis', include=is_latex()}
  mytable(movies$Animation, movies$Action, ci=0.95, prec="%.1f", prec_p="%.4f", latex="rp{4cm}p{4cm}", pct_sign="%")
```


#### Cross table for a continuous variable with statistical testing
For group-wise testing of a continuous variable pass a third variable to mytable, e.g. comparing the movie rating in relation to the genres:
```{r, results='asis', include=is_latex()}
  mytable(movies$Animation, movies$Action, movies$budget, latex=TRUE)

# Change the precision of the numbers and the p-value by adding the properties and setting the scientific format with digits:
  mytable(movies$Animation, movies$Action, movies$budget, prec="%.2e", prec_p="%.2e", latex=TRUE)
  mytable(movies$Action, movies$length<60, movies$budget, prec="%.2e", prec_p="%.2e", latex=TRUE)
```


### characteristics_table

This is a helpful function to compute and display a characteristics table of the study population with separate columns for each cohort, e.g. treatment vs. control. In our example data set we stratify the movie database by `Documentary` which is a binary variable, indicating whether the movie is a documentary or not.
characteristics_table.R will do the job of printing characteristics seperated by a binary or factorial response (genre). P-values are the results of statistical testing comparing both groups: T test and Wilcoxon ranksum test for continous variables with mean and SD or median and quartiles respectively. Statistical testing with categorical data is conducted by Fishers exact test or Chi-Squared-Test (categorical with more than 3 levels). Missing values (NA’s) was omitted for this analysis. Precision of numerials can be set individually for continuous values and p values. Default is one digit after decimal place `prec="%.1f"`, `prec_continuous="%.0f"` and four digits for p values `prec_p="%.4f"`.

```{r, results='asis', include=is_latex()}
# For LaTeX output as a longtable use the following command.
# The compiling of the LaTeX script to a PDF needs the inclusion of the **longtable** in the header.
  characteristics_table(-2, "Documentary", movies[,-1], "col", prec="%.1f", prec_continuous="%.1f", latex="p{1.5cm}p{2cm}rrrr", tablefootnote=FALSE)
```

It also works for categorical variables with more than two levels, such as `genre`, which has six levels specified in movies with a single/unique genre assignment. Statistical testing will change to Chi-Squared-Testing, Kruskal-Wallis rank sum test and One-way analysis of variance (ANOVA). Footnotes will tell you the statistical test behind the p values.
```{r, results='asis', include=is_latex()}
# You may change the footnote labeling as follows:
  characteristics_table(-2, "genre", movies[,-c(1,7:16)], "col", prec="%.1f", prec_continuous="%.1f", prec_p="%.3f", tablefootnote=TRUE,  latex="p{1.2cm}p{1.5cm}p{1.4cm}p{1.4cm}p{1.4cm}p{1.4cm}p{1.4cm}p{1.4cm}rr", fn=c("'", "_", "''", "_", "'''", "_"))

# Or firstly store the results, and use stargazer_long for the subsequent LaTeX output, e.g. with rotation:
  movies$budget = movies$budget/1e6
  s = characteristics_table(-2, "genre", movies[,-c(1,7:16)], "col", prec="%.1f", prec_continuous="%.1f", prec_p="%.3f", tablefootnote=FALSE)
  stargazer_long(s, summary=FALSE, rownames=FALSE, output="p{1.2cm}p{1.5cm}p{1.4cm}p{1.4cm}p{1.4cm}p{1.4cm}p{1.4cm}p{1.4cm}rr", rotate=60)
```


\newpage
## Rapport

<rapport>[http://rapport-package.info/](rapport) is an R package that facilitates creation of reproducible statistical report templates. Once created, rapport templates can be exported to various external formats: HTML, LaTeX, PDF, ODT, DOCX etc.

Let's have short look at the available templates:
```{r}
rapport.ls()
```

Lets try to rebuild the Wilcoxon test results and compare with mytable results:
```{r, results='asis', include=is_latex()}
  #rapport('WilcoxonTest.tpl', data=movies, ...)

  #mytable(movies$Animation, movies$Action, movies$budget, prec="%.2e", prec_p="%.2e", latex=TRUE)
```
... to be continued